Report on the Second Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE2)

### 2015-05-22 WSSSPE2_report/WSSSPE2_review.tex
Update WSSSPE2_review.tex
lots of little changes throughout

\author{Daniel S. Katz$^{(1)}$, Sou-Cheng T. Choi$^{(2)}$, Nancy Wilkins-Diehr$^{(3)}$, Neil Chue Hong$^{(4)}$, 
\\James Howison$^{(5)}$, Miguel de Val-Borro$^{(6)}$, Richard Littauer$^{(7)}$, more TBD %by contributions 
\\{\scriptsize\today \quad  (Due:  $\backsim$ April 30)}} 
%\author{Gabrielle Allen$^{(11)}$,  \\
%Karen Cranston$^{(4)}$, Manish Parashar$^{(5)}$, David Proctor$^{(6)}$, \\
%Matthew Turk$^{(2)}$, Colin C. Venters$^{(7)}$,  Miguel de Val-Borro$^{(10)}$}
%
\thanks{{}$^{(1)}$ Computation Institute, University of Chicago \& Argonne National Laboratory, Chicago, IL, USA; \url{dsk@uchicago.edu}}
%
\thanks{{}$^{(2)}$ NORC at the University of Chicago and Illinois Institute of Technology, Chicago, IL, USA; \url{sctchoi@uchicago.edu}}
%
\thanks{{}$^{(3)}$ University of California-San Diego, San Diego, CA, USA; \url{wilkinsn@sdsc.edu}}
%
\thanks{{}$^{(4)}$ Software Sustainability Institute, University of Edinburgh, Edinburgh, UK; \url{N.ChueHong@software.ac.uk}}
%
\thanks{{}$^{(5)}$ University of Texas at Austin, Austin, TX, USA; \url{jhowison@ischool.utexas.edu}}
%
\thanks{{}$^{(6)}$ Princeton University, Princeton, NJ, USA; \url{todo}}  
%
\thanks{{}$^{(7)}$ TODO; \url{todo}}   
%  
%\thanks{{}$^{(4)}$ National Evolutionary Synthesis Center, Durham, NC, USA}
%
%\thanks{{}$^{(5)}$ Rutgers Discovery Informatics Institute, Rutgers University, New Brunswick, NJ, USA}
%
%\thanks{{}$^{(6)}$ International Consortium of Research Staff Associations, Dublin, Ireland}
%
%\thanks{{}$^{(7)}$ University of Huddersfield, School of Computing and Engineering, Huddersfield, UK}
%
%\thanks{{}$^{(9)}$ Saarland University, Saarbr\"{u}cken, Germany}
%
%\thanks{{}$^{(10)}$ Department of Astrophysical Sciences, Princeton University, Princeton, NJ, USA}
%
%\thanks{{}$^{(11)}$ University of Illinois, Champaign, IL, USA}

@@

Progress in scientific research is dependent on the quality and accessibility of
software at all levels. Hence it is critical to address challenges related to
the development, deployment, maintenance, and overall sustainability of reusable
software as well as education around software practices. These challenges can be
technological, policy based, organizational, and educational, and are of
interest to developers (the software community), users (science disciplines),
software-engineering researchers, and researchers studying the conduct of
science (science of team science, science of organizations, science of science
and innovation policy, and social science communities). The WSSSPE1 workshop
engaged the broad scientific community to identify challenges and best practices
in areas of interest to creating sustainable scientific software. WSSSPE2
invited the community to propose and discuss specific mechanisms to move towards
an imagined future practice for software development and usage in science and
engineering. The workshop included multiple mechanisms for participation,
encouraged team building around solutions, and identified risky solutions with
potentially transformative outcomes. It strongly encouraged participation of
early-career scientists, postdoctoral researchers, and graduate students, with
funds provided to the conference organizers by the Moore Foundation and the
National Science Foundation (NSF), to support the travel of potential
participants who would not otherwise be able to attend, and young participants
and those from underrepresented groups, respectively. These funds allowed
\todo{number} additional participants to attend, and each was offered the chance
to present a lightning talk.

@@

\item \textbf{Daniel Katz: Implementing Transitive Credit with
JSON-LD~\cite{Katz_transitive_credit_poster}.} Science and engineering research
increasingly relies on activities that facilitate research but are not rewarded
or recognized, such as: data sharing; developing common data resources, software
and methodologies; and annotating data and publications. To promote and advance
these activities, we must develop mechanisms for assigning credit, facilitate
the appropriate attribution of research outcomes, devise incentives for
activities that facilitate research, and allocate funds to maximize return on
investment. Katz discussed the issue of assigning credit for both direct and
indirect contributions by using JSON-LD to implement a prototype transitive
credit system.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Exploring Sustainability} \label{sec:exploring}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\todo{Colin to lead writing of this section.}

@@

\todo{short intro to group here}
Six papers were categorized under the theme of Exploring Sustainability. The group included four of the authors from the six papers submitted and an additional number of participants who expressed an interest in the theme at the workshop. Each paper had a different perspective on the concept of sustainability, which ranged from the sustainment of communities to defining sustainability as a first-class, composite non-functional requirement.
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\subsection{Discussion and Actions}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Software Development Experiences} \label{sec:devel}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\todo{Nancy to lead writing of this section}

@@

The group also discussed ``attribution trees,'' an idea put forward by Dan Katz
and Arfon Smith where a chain of attributions can, for example, give credit to
those developing libraries and building blocks that other pieces of software
use. The group considered potential journals and medium to push the application
of this idea. One participant noted that Dryad\footnote{The Dryad Digital
Repository, \url{http://datadryad.org}.} works with journals. If a paper is
accepted, the supporting data must be submitted. We discussed reproducibility as
a component of the journal review process and ``active papers,'' with immediate
links to the data and software. \choinote{Connect the last two sentences to the preceeding materials.}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Credit \& Incentives} \label{sec:credit}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\todo{Dan leading writing of this section}

@@

The following general points were initially discussed: 
\begin{itemize}
\item The FLASH~\cite{flash} project was suggested as an example of how something 
like this has been done.
\item This data could be a file that can be associated with a citation to the 
software, either through use of a DOI for the credit file, or by uploading the 
credit files as associated with the paper.
\item This idea could also be applied to data
\item This data (the credit file) should be part of the metadata that is outside
of paywalls, like citations are now. \choinote{what are ``paywalls''?}
\item It was suggested that there should also be a separate file to track
software dependencies, but this is not the same file. 
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Reproducibility \& Reuse \& Sharing} \label{sec:reproduce}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\todo{Neil OR VOLUNTEER to lead writing of this section}

>>>>

\author{Daniel S. Katz$^{(1)}$, Sou-Cheng T. Choi$^{(2)}$, Nancy Wilkins-Diehr$^{(3)}$, Neil Chue Hong$^{(4)}$, 
\\Colin C. Venters$^{(5)}$, James Howison$^{(6)}$, Frank Seinstra$^{(7)}$, Miguel de Val-Borro$^{(8)}$, Richard Littauer$^{(9)}$, more TBD by contributions}
%\author{Gabrielle Allen$^{(11)}$,  \\
%Karen Cranston$^{(4)}$, Manish Parashar$^{(5)}$, David Proctor$^{(6)}$, \\
%Matthew Turk$^{(2)}$,  Miguel de Val-Borro$^{(10)}$}
%
\thanks{{}$^{(1)}$ Computation Institute, University of Chicago \& Argonne National Laboratory, Chicago, IL, USA; \url{dsk@uchicago.edu}}
%
\thanks{{}$^{(2)}$ NORC at the University of Chicago and Illinois Institute of Technology, Chicago, IL, USA; \url{sctchoi@uchicago.edu}}
%
\thanks{{}$^{(3)}$ University of California-San Diego, San Diego, CA, USA; \url{wilkinsn@sdsc.edu}}
%
\thanks{{}$^{(4)}$ Software Sustainability Institute, University of Edinburgh, Edinburgh, UK; \url{N.ChueHong@software.ac.uk}}
%
\thanks{{}$^{(5)}$ University of Huddersfield, School of Computing and Engineering, Huddersfield, UK}
%
\thanks{{}$^{(6)}$ University of Texas at Austin, Austin, TX, USA; \url{jhowison@ischool.utexas.edu}}
%
\thanks{{}$^{(7)}$ Netherlands eScience Centre, Amsterdam, Netherlands \url{F.Seinstra@esciencecenter.nl}}   
%
\thanks{{}$^{(8)}$ Princeton University, Princeton, NJ, USA; \url{mdevalbo@astro.princeton.edu}}  
%
\thanks{{}$^{(9)}$ MIT, Boston, MA, USA \& University of Saarland, Germany \url{richard.littauer@gmail.com}}   
%  
%\thanks{{}$^{(4)}$ National Evolutionary Synthesis Center, Durham, NC, USA}
%
%\thanks{{}$^{(5)}$ Rutgers Discovery Informatics Institute, Rutgers University, New Brunswick, NJ, USA}
%
%\thanks{{}$^{(6)}$ International Consortium of Research Staff Associations, Dublin, Ireland}
%
%\thanks{{}$^{(9)}$ Saarland University, Saarbr\"{u}cken, Germany}
%
%\thanks{{}$^{(10)}$ Department of Astrophysical Sciences, Princeton University, Princeton, NJ, USA}
%
%\thanks{{}$^{(11)}$ University of Illinois, Champaign, IL, USA}

@@

Progress in scientific research is dependent on the quality and accessibility of
software at all levels. Hence it is critical to address challenges related to
the development, deployment, maintenance, and overall sustainability of reusable
software as well as education around software practices. These challenges can be
technological, policy based, organizational, and educational, and are of
interest to developers (the software community), users (science disciplines),
software-engineering researchers, and researchers studying the conduct of
science (science of team science, science of organizations, science of science
and innovation policy, and social science communities). The WSSSPE1 workshop
engaged the broad scientific community to identify challenges and best practices
in areas of interest to creating sustainable scientific software. WSSSPE2
invited the community to propose and discuss specific mechanisms to move towards
an imagined future practice for software development and usage in science and
engineering. The workshop included multiple mechanisms for participation,
encouraged team building around solutions, and identified risky solutions with
potentially transformative outcomes. It strongly encouraged participation of
early-career scientists, postdoctoral researchers, and graduate students, with
funds provided to the conference organizers by the Moore Foundation and the
National Science Foundation (NSF), to support the travel of potential
participants who would not otherwise be able to attend, and young participants
and those from underrepresented groups, respectively. These funds allowed
17 additional participants to attend, and each was offered the chance
to present a lightning talk.

@@

\item \textbf{Daniel S. Katz: Implementing Transitive Credit with
JSON-LD~\cite{Katz_transitive_credit_poster}.} Science and engineering research
increasingly relies on activities that facilitate research but are not rewarded
or recognized, such as: data sharing; developing common data resources, software
and methodologies; and annotating data and publications. To promote and advance
these activities, we must develop mechanisms for assigning credit, facilitate
the appropriate attribution of research outcomes, devise incentives for
activities that facilitate research, and allocate funds to maximize return on
investment. Katz discussed the issue of assigning credit for both direct and
indirect contributions by using JSON-LD to implement a prototype transitive
credit system.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Exploring Sustainability} \label{sec:exploring}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\todo{Colin led writing of this section.}

@@

Six papers were categorized under the theme of Exploring Sustainability. The group included four of the authors 
from the six papers submitted and an additional number of participants who expressed an interest in the theme at the 
workshop. Each paper had a different perspective on the concept of sustainability, which ranged from the sustainment of 
communities to defining sustainability as a first-class, composite non-functional requirement.
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\subsection{Discussion and Actions}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\subsubsection{Discussion}

Each author was invited to outline the key action from their paper as a potential discussion point for the group; in 
their absence, the group facilitators outlined the actionable points from papers where the authors were not present. 
The key actions from the six papers included recommendations for improving practice in sustainable software 
engineering~\cite{wssspe2_rosada_de_souza, wssspe2_downs}, the development of Software and Infrastructure as a Service 
as a mechanism for fostering sustainable science communities~\cite{wssspe2_patra}, developer incentives for code contributions 
to open source projects~\cite{wssspe2_pierce}, the establishment of a set software engineering principles based in 
scalability, reproducibility, and energy efficiency~\cite{wssspe2_shi}, and applying software architectures as a mechanism 
for architectural-level reasoning about sustainability~\cite{wssspe2_venters}. The group took a position to view 
sustainability from the perspective of addressing the challenges related to the development, deployment, and maintenance of 
reusable software~\cite{XXX}. \katznote{fix cite} The principal focus of the initial discussion considered how to foster cultural change towards 
developing sustainable software in academic environments. It was suggested that a new requirement driven by the agencies 
funding research projects where software was an intrinsic part of enabling the research program would be an additional 
element within the grant proposal to provide a sustainability plan for sustaining the software. This type of initiative would 
provide the necessary incentive and motivation for researchers to consider how to sustain their software beyond the lifetime 
of the project.

\subsubsection{Actions}

The main action to come from the group was a proposal to identify the ten best software engineering practices 
similar to Philip Bourne's approach of Ten Simple Rules~\cite{bourne-ten-simple-rules}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Software Development Experiences} \label{sec:devel}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\todo{Nancy led writing of this section}

@@

The group also discussed ``attribution trees,'' an idea put forward by Dan Katz
and Arfon Smith where a chain of attributions can, for example, give credit to
those developing libraries and building blocks that other pieces of software
use. The group considered potential journals and medium to push the application
of this idea. One participant noted that Dryad\footnote{The Dryad Digital
Repository, \url{http://datadryad.org}.} works with journals. If a paper is
accepted to one of those journals, the supporting data must be submitted to Dryad.
The group also discussed reproducibility as
a component of the journal review process and ``active papers,'' with immediate
links to the data and software.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Credit \& Incentives} \label{sec:credit}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\todo{Dan led writing of this section}

@@

The following general points were initially discussed: 
\begin{itemize}
\item The FLASH~\cite{flash} project was suggested as an example of how something 
like this has been done.
\item This data could be a file that can be associated with a citation to the 
software, either through use of a DOI for the credit file, or by uploading the 
credit files as associated with the paper.
\item This idea could also be applied to data
\item This data (the credit file) should be part of the metadata that are freely
available to those who have not paid for access to the journal, like citations are now in most cases.
\item It was suggested that there should also be a separate file to track
software dependencies, but this is not the same file. 
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Reproducibility \& Reuse \& Sharing} \label{sec:reproduce}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\todo{Neil led writing of this section}
